Who Stays Single? Exploring the Factors Behind Non-Marriage in the United States

### 2024-11-26 paper/paper.qmd
update paper

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Xizi Sun
thanks: "Code and data are available at: https://github.com/LilianS77/US_Marriage."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

@@

# Plot
ggplot(marital_status_df, aes(x = Marital_Status, y = Proportion, fill = Marital_Status)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Proportion of Marital Status Categories",
    x = "Marital Status",
    y = "Proportion"
  ) +
  theme_minimal()
```

@@

# Model

@@

## Results

@@

# Separate pie charts for each gender with percentage labels
ggplot(gender_marital_status, aes(x = "", y = proportion, fill = marital_status)) +
  geom_bar(stat = "identity", width = 1) +
  geom_text(
    aes(label = percentage_label),
    position = position_stack(vjust = 0.5), size = 4, color = "white"
  ) +
  coord_polar("y", start = 0) +
  facet_wrap(~gender) +
  scale_fill_manual(values = c("Not_Married" = "#FF9999", "Married" = "#66B3FF")) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Marital Status"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    strip.text = element_text(size = 14),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```
# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

>>>>

---
title: "Who Stays Single? Exploring the Factors Behind Non-Marriage in the United States"
subtitle: "Low Income and Education Levels Correlate with Higher Propensity to Remain Unmarried"
author: 
  - Xizi Sun
thanks: "Code and data are available at: https://github.com/LilianS77/US_Marriage."
date: today
date-format: long
abstract: "This study examines the socio-demographic determinants of non-marriage in the United States using logistic regression analysis on individual-level data. The findings highlight that individuals with lower income and education levels, particularly below a bachelor’s degree, are significantly more likely to remain unmarried. Additionally, gender and race also play critical roles, with men and certain racial groups, such as African Americans, showing higher unmarried rates. These results underscore structural inequities and socio-economic barriers shaping marital outcomes, offering insights into broader societal patterns and potential areas for intervention."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

Marriage is often regarded as a cornerstone of social and economic stability, shaping the lives of individuals and the broader dynamics of society. However, in recent decades, the United States has experienced a notable shift in marital patterns, with an increasing proportion of the population remaining unmarried. This trend raises essential questions about the underlying factors contributing to the decision not to marry and the societal implications of such changes. Addressing these questions is vital for understanding broader socio-economic inequalities and cultural dynamics.

This paper examines the determinants of non-marriage by analyzing a rich dataset from IPUMS USA [@ipumsusa], which offers harmonized census and survey data spanning demographic, social, and economic variables. Prior studies have largely focused on individual aspects of marital behavior, such as income or education, but few have comprehensively analyzed the intersection of multiple socio-demographic factors. This study bridges that gap by exploring the combined effects of education, income, race, gender, and age on marital outcomes.

Using logistic regression analysis, the results reveal compelling patterns: individuals with higher education levels are more likely to marry, and economic stability, as indicated by higher income, is positively associated with marital status. Moreover, significant disparities are observed across racial and gender groups. African Americans exhibit higher non-marriage rates, while males show slightly elevated probabilities of remaining unmarried compared to females. These findings highlight persistent structural inequalities and cultural nuances shaping marriage in contemporary America.

Understanding the factors influencing non-marriage is crucial for addressing broader societal issues, including economic inequality, cultural shifts, and demographic changes. Marriage is often linked to financial stability, access to social networks, and overall well-being. The disparities uncovered in this study highlight systemic inequities that disproportionately affect certain groups, such as racial minorities and individuals with lower educational attainment. By identifying these disparities, policymakers and social institutions can better target interventions to reduce inequalities, promote inclusive opportunities for stable relationships, and adapt to the changing dynamics of family structures in modern society.

The remainder of this paper is organized as follows. @sec-data describes the dataset, variable selection, and measurement strategies used to examine marital status. @sec-model outlines the logistic regression approach, including the rationale behind model choice, prior specifications, and validation methods. @sec-results section presents key findings, supported by visualizations, and highlights the roles of education, income, race, gender, and age in predicting non-marriage. Finally, @sec-discussion explores the implications of these findings, acknowledges limitations, and proposes directions for future research.

@@

# Plot
ggplot(marital_status_df, aes(x = Marital_Status, y = Proportion, fill = Marital_Status)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Marital Status",
    y = "Proportion"
  ) +
  theme_minimal()
```

@@

# Model {#sec-model}

@@

## Results {#sec-results}

@@

# Separate pie charts for each gender with percentage labels
ggplot(gender_marital_status, aes(x = "", y = proportion, fill = marital_status)) +
  geom_bar(stat = "identity", width = 1) +
  geom_text(
    aes(label = percentage_label),
    position = position_stack(vjust = 0.5), size = 4, color = "white"
  ) +
  coord_polar("y", start = 0) +
  facet_wrap(~gender) +
  scale_fill_manual(values = c("Not_Married" = "#FF9999", "Married" = "#66B3FF")) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Marital Status"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    strip.text = element_text(size = 14),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```
# Discussion {#sec-discussion}

## Determinants of Non-Marriage and Societal Implications

This study delves into the socio-demographic factors that influence marital status, with a specific focus on the probability of remaining unmarried in the United States. By analyzing individual-level microdata through logistic regression, we identified significant predictors such as education, income, race, gender, and age.

For instance, the logistic regression model revealed that individuals with higher education levels, particularly those with a bachelor's degree or more, were significantly less likely to remain unmarried compared to those with lower education levels. Education, therefore, serves as a strong stabilizer in marital outcomes, providing individuals with economic resources and social capital that facilitate marriage. Simultaneously, the analysis underscored the heightened likelihood of remaining unmarried among African Americans, reflecting deeper structural inequities that impact marital patterns. Gender also emerged as a significant factor, with males exhibiting a slightly higher probability of being unmarried compared to females.

These findings, grounded in data and supported by statistical rigor, affirm prior literature while adding layers of nuance. They emphasize how individual socio-demographic traits intersect with broader societal structures to shape marital outcomes.

## Insights into Education, Income, and Marriage Patterns

One of the most compelling findings of this study is the pronounced impact of education on marital status. Individuals with higher education levels—those with bachelor's degrees or above—demonstrated significantly lower rates of non-marriage. This is evident from the analysis showing that only 16% of individuals with education above a bachelor's degree were unmarried, compared to 32% of those with "Some College" education. The logistic regression coefficients also reinforced this relationship, with those holding higher degrees being far less likely to remain unmarried (e.g., coefficient for "Below High School" = 0.48).

This finding aligns with existing theories that higher education fosters economic stability, enhances social capital, and provides a greater sense of agency in life decisions, all of which are conducive to marriage. However, the results also highlighted that income, while less strongly predictive, plays a complementary role. The income distribution plot for unmarried individuals revealed clustering at lower income brackets, with the majority earning below $40,000 annually. This suggests that financial security continues to be an essential consideration in marital decisions, albeit to a lesser extent than education.

These patterns point to the interplay between socio-economic resources and marital outcomes, emphasizing how disparities in access to education and income perpetuate inequalities in family formation.

## Gender and Race in Marital Trends

The analysis revealed striking disparities in marital trends across gender and racial lines. Gender-wise, men exhibited a higher likelihood of remaining unmarried compared to women, with 30% of men being unmarried versus 26% of women. This aligns with cultural norms and gendered expectations, where women are often perceived as more closely tied to family roles and marriage. Moreover, women may face greater societal pressure to marry, which could explain their lower rates of remaining unmarried.

Racial disparities in marital status were even more pronounced. For example, African Americans exhibited the highest rates of non-marriage (43%), significantly higher than those for Asian Americans (30%) and White individuals (24%). These patterns likely reflect a combination of structural inequalities, cultural factors, and historical contexts. Economic disparities, higher unemployment rates, and systemic barriers to wealth accumulation among African Americans may contribute to these higher non-marriage rates. Conversely, Asian Americans, who exhibited the lowest rates of remaining unmarried, may benefit from cultural norms that strongly prioritize marriage and family cohesion.

These findings underscore how structural inequities and cultural norms interact to shape marital trends. They also highlight the need for policies aimed at addressing systemic barriers and promoting equality in socio-economic outcomes to reduce disparities in family formation.

## Weaknesses and Next Steps

While this study provides critical insights into the predictors of marital status, it is not without limitations. The reliance on cross-sectional data, for example, restricts the ability to establish causal relationships. While the findings suggest strong associations between education, income, race, gender, and marital outcomes, longitudinal data could better capture how these factors interact over time to influence marriage decisions.

Another limitation lies in the dataset itself, which does not include qualitative factors such as cultural attitudes, personal preferences, or psychological traits. These dimensions are critical for understanding the broader context of marital decisions. For instance, while income and education are robust predictors, societal expectations and individual life goals may play equally important roles that remain unquantified in this analysis.

Future research should aim to integrate mixed-method approaches, combining quantitative rigor with qualitative depth, to provide a holistic view of marital trends. Additionally, while this study focused exclusively on the United States, extending the analysis to include international data could uncover global patterns and cultural nuances. For example, exploring how marriage trends differ in countries with varying economic systems, religious practices, and gender norms would provide valuable comparative insights.

## Conclusion

This study contributes to the understanding of non-marriage determinants by highlighting the interplay between education, income, race, gender, and age. Education emerged as the most significant predictor, with higher levels strongly associated with lower non-marriage rates. Income, while secondary, still played a notable role, emphasizing the importance of economic stability in marital decisions. Gendered expectations and racial disparities further revealed how societal structures and cultural norms influence marriage patterns.

The findings highlight the socio-economic inequities that shape family formation and underscore the importance of addressing structural barriers to promote marital stability. Future research should continue to explore these dynamics, leveraging longitudinal data and integrating qualitative perspectives to deepen our understanding of marriage and its evolving role in contemporary society.
